Manuscript for Foodborne-Outbreak Project

### 2023-04-24 products/manuscript/Manuscript.qmd
add vip to analysis and manuscript

---
title: "Manuscript for foodborne-outbreak Project"
subtitle: ""
author: Andreas Handel
date: "`r Sys.Date()`"
format:
  docx:
    toc: false
    number-sections: true
    highlight-style: github
bibliography: ../references.bib
csl: ../apa.csl
editor: 
  markdown: 
    wrap: sentence
---

>>>>

---
title: "Manuscript for Foodborne-Outbreak Project"
# subtitle: ""
author: "Weifan Wu and Kailin Chen"
# date: "`r Sys.Date()`"
format: 
  docx
#    toc: false
#    number-sections: true
#    highlight-style: github
# bibliography: ../references.bib
# csl: ../apa.csl
# editor: 
#  markdown: 
#    wrap: sentence
---

### 2023-04-24 products/manuscript/Manuscript2.qmd
add vip to analysis and manuscript

```{r echo = FALSE, warning = FALSE, paged.print = TRUE}
# Load Libraries
library(here)
library(knitr)
library(tidyverse)
library(broom)
library(ggtext)
library(flextable)
```

@@

After applying the aforementioned data processing method to the raw data, a total of 3985 observations were obtained in the training data set. The illnesses caused by a single outbreak ranged from `r min (training_data_final$Illnesses)` to `r max (training_data_final$Illnesses)` while resulted hospitalizations varied from `r min (training_data_final$Hospitalizations)` to `r max (training_data_final$Hospitalizations)` from 1998 to 2015. In the United States, Multi-state outbreaks were the most commonly associated with total hospitalizations (N= `r max (state_hosp$all_hospitalizations)`) caused by foodborne illnesses, followed by California (N=`r state_hosp$all_hospitalizations[2]`), Ohio (N=`r state_hosp$all_hospitalizations[3]`), and New York (n=`r state_hosp$all_hospitalizations[4]`). Poultry, fruits, beef, and pork were identified as the most frequent food vehicles linked to mass hospitalizations (involving over 500 individuals) due to foodborne illnesses (Figure 2). Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2015, as shown in Table 2. Figure 1 unveiled patterns concerning the seasonality of foodborne illness outbreaks. Hospitalizations due to foodborne illnesses peaked in July, while reaching their lowest point between December and February. Intriguingly, bacteria is the predominant etiologic agent that causes hospitalization (Figure 3). Among these bacteria, the genus *Salmonella* is the most frequently identified etiologic agent responsible for foodborne illness-related hospitalizations.Most commonly implicated food vehicles for *Salmonella* include fruits, pork, and poultry (Figure 5). In contrast, *E. coli* and *Clostridium*, two other bacterial species associated with foodborne hospitalizations, were most frequently linked to vegetables as the primary food carrier. Restaurant and private home were locations where the majority of foodborne hospitalizations occurred (Figure 6). The testing data set revealed a number of trends that were consistent with those found in the training data set. For instance, *Salmonella* continued to be the most common bacterial pathogen responsible for foodborne illnesses, and restaurants remained the most frequently implicated exposure setting (Figure 7 and Figure 8). However, some shifts were also observed between the two data sets. While Table 2 had previously shown salad to be the leading food carrier of foodborne illnesses, the testing data set revealed that poultry was actually the most common source of illness between 2009 and 2020 (Figure 9). In addition, the testing data set revealed that *E. coli* had replaced the previously identified second most common bacterial pathogen *Clostridium*, causing a significant number of illnesses between 2009 and 2020 (Figure 8).

@@

Lastly, the random forest model outperformed all other models. It is also interesting to note that bagging resulted in a higher RMSE than random forest each time despite using all predictors in its splitting, which should have greatly reduced variance. However, random forest models take a random subset of the 10 predictors for splitting; average variance is thus reduced from high to moderate--meaning that random forest should have performed worse. Additional future investigation is suggested to uncover why.

@@

![Figure 3. Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png) ![Figure 4. Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png) ![Figure 5. Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png) ![Figure 5. Leading locations that result in hospitalizations](images/Location_v_hospitalization.png) ![Figure 6. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png) ![Figure 7. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png) ![Figure 8. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png) ![Figure 9. Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)

>>>>

```{r include=FALSE,echo = FALSE, warning = FALSE}
# Load Libraries
library(here)
library(knitr)
library(tidyverse)
library(broom)
library(ggtext)
library(flextable)
```

@@

After applying the aforementioned data processing method to the raw data, a total of 3985 observations were obtained in the training data set. The illnesses caused by a single outbreak ranged from `r min (training_data_final$Illnesses)` to `r max (training_data_final$Illnesses)` while resulted hospitalizations varied from `r min (training_data_final$Hospitalizations)` to `r max (training_data_final$Hospitalizations)` from 1998 to 2015. In the United States, Multi-state outbreaks were the most commonly associated with total hospitalizations (N= `r max (state_hosp$all_hospitalizations)`) caused by foodborne illnesses, followed by California (N=`r state_hosp$all_hospitalizations[2]`), Ohio (N=`r state_hosp$all_hospitalizations[3]`), and New York (n=`r state_hosp$all_hospitalizations[4]`). Poultry, fruits, beef, and pork were identified as the most frequent food vehicles linked to mass hospitalizations (involving over 500 individuals) due to foodborne illnesses (Figure 2). Nevertheless, despite poultry being the top food vehicle, salad was identified as the primary culprit for causing the most foodborne illnesses from 1998 to 2015, as shown in Table 2.

Figure 1 unveiled patterns concerning the seasonality of foodborne illness outbreaks. Hospitalizations due to foodborne illnesses peaked in July, while reaching their lowest point between December and February. Intriguingly, bacteria is the predominant etiologic agent that causes hospitalization (Figure 3). Among these bacteria, the genus *Salmonella* is the most frequently identified etiologic agent responsible for foodborne illness-related hospitalizations.Most commonly implicated food vehicles for *Salmonella* include fruits, pork, and poultry (Figure 5). In contrast, *E. coli* and *Clostridium*, two other bacterial species associated with foodborne hospitalizations, were most frequently linked to vegetables as the primary food carrier. Restaurant and private home were locations where the majority of foodborne hospitalizations occurred (Figure 6). The testing data set revealed a number of trends that were consistent with those found in the training data set. For instance, *Salmonella* continued to be the most common bacterial pathogen responsible for foodborne illnesses, and restaurants remained the most frequently implicated exposure setting (Figure 7 and Figure 8). However, some shifts were also observed between the two data sets. While Table 2 had previously shown salad to be the leading food carrier of foodborne illnesses, the testing data set revealed that poultry was actually the most common source of illness between 2009 and 2020 (Figure 9). In addition, the testing data set revealed that *E. coli* had replaced the previously identified second most common bacterial pathogen *Clostridium*, causing a significant number of illnesses between 2009 and 2020 (Figure 8).

@@

Lastly, the random forest model outperformed all other models in minimizing RMSE. Our findings reveal that the 'Multistate' variable holds the highest importance in determining hospitalization predictions, followed by 'Year', 'Month', and 'IFSAC Category_Fruits'. 'Location' also plays a significant role in the model's predictive power (Figure 10). These results suggest that geographical factors, such as multistate outbreaks, and temporal factors, such as year and month, play a crucial role in predicting hospitalizations. Furthermore, the specific IFSAC category, particularly 'Fruits', and the presence of *Salmonella* contribute significantly to the model's performance, indicating that certain food categories and bacterial etiologies may have a higher association with hospitalization rates.

It is also interesting to note that bagging resulted in a higher RMSE than random forest each time despite using all predictors in its splitting, which should have greatly reduced variance. However, random forest models take a random subset of the 10 predictors for splitting; average variance is thus reduced from high to moderate--meaning that random forest should have performed worse. Additional future investigation is suggested to uncover why.

@@

![Figure 3. Pathogen types and Hospitalizations.](images/pathogen_type_v_hospitalizations.png) ![Figure 4. Common bacterial etiology causing hospitalizations from 1998-2015.](images/bacterial_etiology_v_hospitalization_frequency.png) ![Figure 5. Top 3 bacterial pathogens and their common food carrier month from 1998-2015.](images/top3_bacterial_etiology_v_food_hospitalization_frequency.png) ![Figure 5. Leading locations that result in hospitalizations](images/Location_v_hospitalization.png) ![Figure 6. Illness over time by pathogen type (2009-2020).](images/time_v_illness_pathogen_type.png) ![Figure 7. Illnesses by outbreak location (2009-2020).](images/Location_v_Illness.png) ![Figure 8. Bacterial etiology by outbreak frequency (2009-2020).](images/bacterial_etiology_v_outbreak_frequency.png) ![Figure 9. Food by outbreak frequency (2009-2020).](images/food_type_v_illness_frequency.png)
![Figure 10. Variable importance in determining hospitalization predictions using random forest.](images/variable importance in predicting hospitalizations by random forest model.png)
